\framework: Morphology-based Entity and Relational Entity Extraction Framework for Arabic

### 2018-04-20 tex/merf/TALMajor/methodology.tex
minor fixes

After relation extraction, we are interested to relate 
entities that express the same concept.
\framework provides the extended synonym feature of second order as 
a default cross-reference relation ($Syn^2$).
In Figure~\ref{fig:intromotiv}, triggering this feature creates the 
edge labels with \cci{isSyn} 
between the nodes \cci{Khalifa Tower} and \cci{The building}.

>>>>

After relation extraction, we are interested to relate 
entities that express the same concept.
\framework provides the extended synonym feature of second order as 
a default cross-reference relation ($Syn^2$).
In Figure~\ref{fig:intromotiv}, triggering this feature creates the 
edge labeled with \cci{isSyn} 
between the nodes \cci{Khalifa Tower} and \cci{The building}.

### 2018-04-20 tex/merf/TALMajor/results.tex
minor fixes

In this section we evaluate \framework with four case studies. 
We perform a survey like evaluation where developers manually 
built task specific information extraction tools for the case studies 
and other developers built equivalent \framework tools. 
The aim of the comparison is to showcase that \framework enables 
fast development of linguistic applications with similar accuracy 
and a reasonable affordable overhead in computational time. 
We report development time,
size of developed code versus size of grammar,
running time, and 
precision-recall as metrics
of cost, complexity, overhead, and accuracy,
respectively.

We survey three case studies from the literature: 
(1) narrator chain, (2) temporal entity, and (3) genealogy entity extraction 
tasks and we use the reported development time for the task specific 
techniques proposed in
ANGE~\cite{ZaMaFlairs2012HadithBio}, 
ATEEMA~\cite{ZaMa2012IJCLATime},  and
GENTREE~\cite{ZaMaHaCicling2012Entity}, respectively. 
%
We also compare a \framework number normalization task to 
a task specific implementation.

We evaluated ANGE with
Musnad Ahmad, a hadith book, where we constructed an annotated 
golden reference containing 1,865 words.
We evaluated ATEEMA 
with articles from issues of  the Lebanese 
Al-Akhbar newspaper where we constructed 
an annotated golden reference containing 1,677 words. 
For the genealogical tree extraction we used 
an extract from the Genesis biblical text with 1,227 words.
Finally, we used an annotated article from the Lebanese Assafir
newspaper with 1,399 words to evaluate the NUMNORM case study
\footnote{available at \url{http://www.assafir.com} and 
\url{http://www.al-akhbar.com}.}. 
%
In the online appendix%
~\footnote{available at ~\url{-omitted-for-anonymity-}}%http://webfea.fea.aub.edu.lb/fadi/pdfs/merfappendix.pdf
, we report on eight additional \framework case studies.
\major{
Manual annotators inspected the outcome and provided corrections where tools 
made mistakes.
The corrections form the manual gold annotation that we compared against.
}

>>>>

\major{
In this section we evaluate \framework with four case studies. 
We perform a survey like evaluation where developers manually 
built task specific information extraction tools for the case studies 
and other developers built equivalent \framework tools. 
The aim of the comparison is to showcase that \framework enables 
fast development of linguistic applications with similar accuracy 
and a reasonable affordable overhead in computational time. 
We report development time,
size of developed code versus size of grammar,
running time, and 
precision-recall as metrics
of cost, complexity, overhead, and accuracy,
respectively.

We survey three case studies from the literature: 
(1) narrator chain, (2) temporal entity, and (3) genealogy entity extraction 
tasks, and we use the reported development time for the task specific 
techniques proposed in
ANGE~\cite{ZaMaFlairs2012HadithBio}, 
ATEEMA~\cite{ZaMa2012IJCLATime},  and
GENTREE~\cite{ZaMaHaCicling2012Entity}, respectively. 
%
We also compare a \framework number normalization task to 
a task specific implementation.

We evaluated ANGE with
Musnad Ahmad, a hadith book, where we constructed an annotated 
golden reference containing 1,865 words.
We evaluated ATEEMA 
with articles from issues of  the Lebanese 
Al-Akhbar newspaper where we constructed 
an annotated golden reference containing 1,677 words. 
For the genealogical tree extraction we used 
an extract from the Genesis biblical text with 1,227 words.
Finally, we used an annotated article from the Lebanese Assafir
newspaper with 1,399 words to evaluate the NUMNORM case study
\footnote{available at \url{http://www.assafir.com} and 
\url{http://www.al-akhbar.com}.}. 
%
In the online appendix%
~\footnote{available at ~\url{-omitted-for-anonymity-}}%http://webfea.fea.aub.edu.lb/fadi/pdfs/merfappendix.pdf
, we report on eight additional \framework case studies.
Manual annotators inspected the outcome and provided corrections where tools 
made mistakes.
The corrections form the manual gold annotation that we compared against.
}

### 2018-04-20 tex/merf/TALMajor/sarf.tex
done till 3.19
minor fixes

%~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
%{\bf Morphological analyzer (Sarf).~}
\major{
Morphological analysis is key to Arabic 
NLP
due its to the exceptional degree of ambiguity in writing, 
the rich morphology, and the complex word derivation 
system~\cite{arabicmorph,shahrour2016camelparser,pasha2014madamira}.
}
Short vowels, also known as diacritics, are typically omitted in Arabic text
and inferred by readers~\cite{habash2006arabic}. 
For example, the word \RL{bn}%
can be interpreted as \RL{bon} {\tt ``coffee''} with a {\em damma} diacritic on the 
letter \utfrl{بـ} or 
\vocalize \RL{bin} (son of) with a 
{\em kasra} diacritic on the letter \utfrl{بـ}.

@@

We denote by 
\ss,
\pp,
\xx,
\PP,
\GG, and 
\AC, the set of 
all stems,
prefixes,
suffixes,
POS,
gloss, 
and user defined category tags, respectively. 
Let $T=\langle t_1,t_2,\ldots,t_M\rangle$ be a set of Arabic words denoting the 
text documents. 
\framework uses Sarf to compute a set of morphological solutions $M(t)=\{m_1,m_2,\ldots,m_N\}$
for each word $t\in T$. 
%
\major{
Each morphological solution $m\in M(t)$ is a tuple of the form 
$\langle p,s,x,P,G,C\rangle \in \pp \times \ss \times \xx \times \PP \times \GG \times \AC$ 
where $p=p_1 \ldots p_{|p|}$, $x=x_1 \ldots x_{|x|}$, 
$P=P_{p_1}\ldots P_{p_{|p|}} P_s P_{x_1}\ldots P_{x_{|x|}}$,
$G=G_{p_1}\ldots G_{p_{|p|}} G_s G_{x_1}\ldots G_{x_{|x|}}$, and
$C=C_{p_1} \ldots C_{p_{|p|}} C_s C_{x_1}\ldots C_{x_{|x|}}$. 
$P_{p_i}, G_{p_i},$ and $C_{p_i},1\le i \le |p|$ are the POS, Gloss and 
category tags of prefix $p_i$. 
$P_{x_j}, G_{x_j},$ and $C_{x_j},1\le j \le |x|$ are the POS, Gloss and 
category tags of suffix $x_i$. 
$P_s, G_s,$ and $C_s$ are 
$P_s, G_s,$ and $C_s$ are the POS, Gloss and 
category tags of stem $s$. 
%
Intuitively, 
$p,x,P,G$ and $C$ are concatenations of prefix, suffix, POS, Gloss and
Category values, respectively.
}

%\setarab
\begin{table}[tb!]
  \centering
  \caption{Sample solution vector for \utfrl{فَسَيَأْكُلها}.}
  \resizebox{0.7\columnwidth}{!}{
    \begin{tabular}{|r|c|c|c|c|c|}
          & \multicolumn{3}{c|}{\textbf{Prefixes}} & \textbf{Stem} & \textbf{Suffix} \\
    \textbf{Data} & \utfrl{فَ} & \utfrl{سَ} & \utfrl{يَ} & \utfrl{أْكُل} & \utfrl{ها} \\
    \textbf{POS} & CONJ+ & FUT+ & IV3MS+ & VERB\_IMPERFECT & IVSUFF\_DO:3FS \\
    \textbf{Gloss} & and/so & will & he/it & eat/consume & it/them/her \\
    \textbf{index} & \multicolumn{3}{c|}{10} & 13 & 16 \\
    \textbf{length} & \multicolumn{3}{c|}{3} & 3 & 2 \\
    \end{tabular}%
    }
  \label{tab:samplesolution}%
\end{table}%

Table~\ref{tab:samplesolution} shows the morphological analysis
of the word \notrutfrl{فَسَيَأْكُلها}. 
The word is composed of the prefix morphemes 
\utfrl{فَ}, \utfrl{سَ}, and \utfrl{يَ}, followed by the 
stem \utfrl{أْكُل}, and then followed by the suffix morpheme
\utfrl{ها}. 
Each morpheme is associated with a number of morphological features.
The {\tt CONJ},
{\tt FUT}, 
{\tt IV3MS} 
{\tt VERB\_IMPERFECT}, and 
{\tt IVSUFF\_DO:3FS} POS tags indicate
conjunction, 
future, 
third person masculine singular subject pronoun,
an imperfect verb, and 
and a third person feminine singular object pronoun, respectively.
The POS and gloss notations follow the Buckwalter notation~\cite{Buckwalter:02}.

>>>>

%~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
%{\bf Morphological analyzer (Sarf).~}
\major{
Morphological analysis is key to Arabic 
NLP
due to the exceptional degree of ambiguity in writing, 
the rich morphology, and the complex word derivation 
system~\cite{arabicmorph,shahrour2016camelparser,pasha2014madamira}.
}
Short vowels, also known as diacritics, are typically omitted in Arabic text
and inferred by readers~\cite{habash2006arabic}. 
For example, the word \RL{bn}%
can be interpreted as \RL{bon} {\tt ``coffee''} with a {\em damma} diacritic on the 
letter \utfrl{بـ} or 
\vocalize \RL{bin} (son of) with a 
{\em kasra} diacritic on the letter \utfrl{بـ}.

@@

We denote by 
\ss,
\pp,
\xx,
\PP,
\GG, and 
\AC, the set of 
all stems,
prefixes,
suffixes,
POS,
gloss, 
and user defined category tags, respectively. 
Let $T=\langle t_1,t_2,\ldots,t_M\rangle$ be a set of Arabic words denoting the 
text documents. 
\framework uses Sarf to compute a set of morphological solutions $M(t)=\{m_1,m_2,\ldots,m_N\}$
for each word $t\in T$. 
%
\major{
Each morphological solution $m\in M(t)$ is a tuple of the form 
$\langle p,s,x,P,G,C\rangle \in \pp \times \ss \times \xx \times \PP \times \GG \times \AC$ 
where $p=p_1 \ldots p_{|p|}$, $x=x_1 \ldots x_{|x|}$, 
$P=P_{p_1}\ldots P_{p_{|p|}} P_s P_{x_1}\ldots P_{x_{|x|}}$,
$G=G_{p_1}\ldots G_{p_{|p|}} G_s G_{x_1}\ldots G_{x_{|x|}}$, and
$C=C_{p_1} \ldots C_{p_{|p|}} C_s C_{x_1}\ldots C_{x_{|x|}}$. 
$P_{p_i}, G_{p_i},$ and $C_{p_i},1\le i \le |p|$ are the POS, Gloss and 
category tags of prefix $p_i$. 
$P_{x_j}, G_{x_j},$ and $C_{x_j},1\le j \le |x|$ are the POS, Gloss and 
category tags of suffix $x_i$. 
$P_s, G_s,$ and $C_s$ are the POS, Gloss and 
category tags of stem $s$. 
%
Intuitively, 
$p,x,P,G$ and $C$ are concatenations of prefix, suffix, POS, Gloss and
Category values, respectively.
}

%\setarab
\begin{table}[tb!]
  \centering
  \caption{Sample solution vector for \utfrl{فَسَيَأْكُلها}.}
  \label{tab:samplesolution}%
  \resizebox{0.7\columnwidth}{!}{
    \begin{tabular}{|r|c|c|c|c|c|}
          & \multicolumn{3}{c|}{\textbf{Prefixes}} & \textbf{Stem} & \textbf{Suffix} \\
    \textbf{Data} & \utfrl{فَ} & \utfrl{سَ} & \utfrl{يَ} & \utfrl{أْكُل} & \utfrl{ها} \\
    \textbf{POS} & CONJ+ & FUT+ & IV3MS+ & VERB\_IMPERFECT & IVSUFF\_DO:3FS \\
    \textbf{Gloss} & and/so & will & he/it & eat/consume & it/them/her \\
    \textbf{index} & \multicolumn{3}{c|}{10} & 13 & 16 \\
    \textbf{length} & \multicolumn{3}{c|}{3} & 3 & 2 \\
    \end{tabular}%
    }
  \vspace{-2em}
\end{table}%

Table~\ref{tab:samplesolution} shows the morphological analysis
of the word \notrutfrl{فَسَيَأْكُلها}. 
The word is composed of the prefix morphemes 
\utfrl{فَ}, \utfrl{سَ}, and \utfrl{يَ}, followed by the 
stem \utfrl{أْكُل}, and then followed by the suffix morpheme
\utfrl{ها}. 
Each morpheme is associated with a number of morphological features.
The {\tt CONJ},
{\tt FUT}, 
{\tt IV3MS} 
{\tt VERB\_IMPERFECT}, and 
{\tt IVSUFF\_DO:3FS} POS tags indicate
conjunction, 
future, 
third person masculine singular subject pronoun,
an imperfect verb, and 
a third person feminine singular object pronoun, respectively.
The POS and gloss notations follow the Buckwalter notation~\cite{Buckwalter:02}.

### 2018-04-20 tex/merf/TALMajor/tool.tex
minor fixes

Given a word $w\in L$, 
$Sy(w)=\{u\mid u\in \ss \land\exists s\in \gamma(w)\land~\alpha(u)\cap\alpha(s)\neq\emptyset\}$
is the set of Arabic stems 
directly related to $w$ through the gloss map.
%
Let $Sy^{i}(w)$ denote stems related to $w$ using the gloss map of order $i$ recursively such that
$Sy^{1}(w) = Sy(w)$ and
$Sy^{i+1}(w)=\{u\mid u\in S\land\exists s\in Sy^{i}(w)\land~\alpha(u)\cap\alpha(s)\neq\phi\}$.
Formally, $Syn^k(w) = \bigcup\limits_{i=1}^{k} Sy^{i}(w)$ for $i\in[1 \ldots k]$.
%
\major{
The example in Figure~\ref{fig:introsynEx} illustrates the computation.
Let $w$ denote an input Arabic word \RL{mA'}, which has the gloss \cci{water}, i.e. \cci{water} $\in\alpha(w)$.
$w$ shares this gloss with the stem \RL{n.d.h}, denoted $s_1$, i.e. $s_1\in Sy^{1}(w)$.
%The input word \RL{mA'} shares the gloss \RL{n.d.h} through the gloss intersection \cci{water}. 
Next, the stem \RL{r^s^s}, denoted $s_2$, shares the gloss \cci{spray} with $s_1$, i.e. $s_2\in Sy^{1}(s1)\subset Sy^{2}(w)$.
Therefore, $Syn^2(w)$ relates the words \RL{mA'} and \RL{r^s^s}.
}

@@

\major{
Consider the task we discussed in the introduction (Figure~\ref{fig:intromotiv}) 
and recall that we are interested in identifying names of people, names of places, relative positions, and numerical terms.
The table in Figure~\ref{fig:taskMBF} presents the defined formulae.
The user denotes the ``name of person'' entities with formula $N$ 
which requires the {\em category} feature in the morphological solution of a word to be {\tt Name\_of \_Person}.
The entities $n_1$, $n_2$, and $n_3$ are matches of the formula $N$ in the text.
Similarly, the user specifies formula $P$ to denote ``name of place'' entities. 
The user specifies formula $R$ to denote ``relative position'' entities, 
and defines it as a disjunction of terms that check for 
solutions matching stems such as \RL{qrb} (near) and \RL{fy} (in).
%requires the stem feature of a word to belong to a selected list of stems containing \RL{fy} and \RL{qrb}. a
Similarly, $U$ denotes numerical terms and is a disjunction of constraints 
requiring the stem feature to belong to a set of stems such as
\RL{'wl}(first), \RL{_tAny}(second), \dots \RL{`A^sr}(tenth).
}

@@

A relation is defined by the user as a tuple 
$\langle e_1,e_2,r\rangle$ where 
$e_1,e_2,$ and $r$ are identifiers associated with 
sub-expressions of an expression $f$.
Matches of the relation are a set of labeled binary edges
where matches of $e_1$ and $e_2$ are the source and destination nodes
and matches of $r$ are the edge labels.
We denote $\ldbrack \langle e_1,e_2,r\rangle \rdbrack$ to be the set of matches of the corresponding relation, 
and we refer to them as relational entities.

\major{
We are interested in constructing the relational entity graph in Figure~\ref{fig:intromotiv}.
Let $e_1,o_1,r,o_2,$ and $e_2$ be identifiers to the sub-expressions 
$(P|N)+$,$O?$,$R$,$O\wedge 2$, and $(P|N|U)+$, respectively.
%We select $(P|N)+$ to denote an entity as it captures a composite place name such as \cci{Khalifa tower}.
We define the semantic relations $\langle e_1, e_2, r\rangle$, $\langle r, e_1, o_1\rangle$, and $\langle r,e_2,o_2\rangle$.
The matches to $e_1$, $r$, $o_2$, and $e_2$ in match 1 (Fig.~\ref{fig:taskMRE}) are \RL{brj _hlyfT} (Khalifa Tower), \RL{bAlqrb} (next), \RL{mn} (to), and \RL{AltqA.t` Al-'wl} (first intersection).
Note that there is no match to the optional $O$ formula in match 1.
Similarly, the matches to $e_1$, $o_1$, $r$, $o_2$, and $e_2$ in the second \major{matching parse tree} are
\RL{dby mwl} (Dubai Mall), \RL{`l_A} (is located), \RL{mqrbT} (near), \RL{mn h_dA} (this), and \RL{Almbn_A} (building), respectively.
}

\major{
Relation $\langle e_1,e_2,r\rangle$ creates the 
edge labeled \cci{next to} between \cci{Khalifa tower} and 
\cci{intersection 1} nodes from match 1, and the 
edge labeled \cci{near} between \cci{Dubai Mall} and \cci{the building} nodes from match 2.
Relation $\langle r,e_1,o_1\rangle$ creates the edge labeled \cci{prep} 
between \cci{Dubai Mall} and \cci{near} nodes from match 2. 
Relation $\langle r,e_2,o_2\rangle$ creates the edge labeled \cci{from} 
between \cci{intersection 1} and \cci{next to} nodes in 
match 1, and the 
edge labeled \cci{from this} between \cci{near} and \cci{the building} 
nodes in match 2.
}

@@

For each word $t_i\in T, 0\le i < |T|$.
\framework computes a Boolean value 
for all MBFs. 
\major{
For example, \RL{brj} matches MBF $P$. 
}
Then, it computes the set of MBF tags
$R_i=\{(t_i,tt)| tt=\langle l,f,d\rangle \wedge
f~\mathit{is~an~MBF} \wedge f(t_i)\} \subseteq T \times {\cal T}$
which tags a word $t_i$ with $\mathit{tt}$ 
iff the MBF $f$ associated with
tag type $\mathit{tt}$ is true for $t_i$. 
The MBF evaluation results in a sequence of tag sets 
$\langle R_0, R_1, \ldots, R_{n-1}\rangle$.
If a word $t_i$ has no tag type match, 
its tag set $R_i$ is by default the singleton $O=\{\mathit{NONE}\}$.
\major{
For example, $R$ for the text in Figure~\ref{f:overview} follows 
$\{ \{NONE\}, \{NONE\}, \{NONE\}, \{NONE\}, \{($\RL{brj}$,P)\}, \{($\RL{xlyfT}$,N)\},\ldots\}$. 
}

@@

\framework simulates the generated NFA over the sequence of tag sets matching the MBF formulae.
%A simulation match $m$ of an expression $f$ is a tree where the root is the MRE expression, the internal nodes are the MRE and MBF operations, and the leaves are matches of the MAT terms of $f$.
A simulation match $m$ of an expression $f$ is a tree where the root spans the expression, the internal nodes are roots to sub-expressions of $f$, and the leaves are matches of the MBF formulae of $f$, e.g. Fig.~\ref{fig:taskMRE}.
The sequence of leaf matches forms a vector of tags $\langle r_k,r_{k+1},\dots,r_j\rangle$ 
corresponding to the text sequence 
$\langle t_k,t_{k+1},\dots,t_j\rangle$ where $r_{\ell}\in R_{\ell},0\le k\le \ell \le j < n$.
%
%If the simulation has a match $\langle r_m,r_{m+1},\dots,r_n\rangle$ 
%where $0\le m\le n$, 
%the next simulation starts at $R_{n+1}$. 
%This disallows overlap of matches for the same MRE. 
%
%In case the NFA simulation has no match, 
%the next simulation starts at $R_{m+1}$. 
If we have more than one match for an expression, 
%starting at $R_k$ where $0\le k\le n$, 
\framework returns the longest.
%Figure~\ref{fig:taskMRE} shows two match trees of $\mathit{dir}$ extracted from the text of Figure~\ref{fig:intromotiv}.
%\RL{dby} and \RL{mwl} are leaf nodes referring to name of place tags ($P$).
%The $+$, $|$, and $?$ MRE operations are internal nodes.

@@

%\begin{figure}[tb!]
%{ \relsize{-1}
%\begin{framed}
%\begin{multicols}{2}
%\begin{itemize}
%\item Annotated Expression: \\
%\\
%{ \relsize{-2.5}
%%\begin{itemize}
%%  \item 
%  $\stackrel{e_1}{(P|N)+}~\stackrel{o_1}{\mathit{O}?}~\stackrel{r}{R}~\stackrel{o_2}{\mathit{O^{\wedge}2}}~\stackrel{e_2}{(P|N|U)+}$
%%\end{itemize}
%}
%\end{itemize}
%\begin{itemize}
%\item User defined semantic \\relations:
%\begin{itemize}
%\item $\langle e_1,e_2,r\rangle$
%\item $\langle r,e_1,o_1\rangle$
%\item $\langle r,e_2,o_2\rangle$
%\end{itemize}
%\end{itemize}
%\columnbreak
%\setcode{utf8}
%\transfalse
%\resizebox{0.9\columnwidth}{!}{
%	\relsize{+1.5} \input{figures/ergraphEx.tex}
%}
%\transtrue
%\setcode{standard}
%\end{multicols}
%\end{framed}
%}
%\caption{User-defined semantic relation example}
%\label{fig:srEx}
%\end{figure}
%
Next, \framework computes the relational entities corresponding to each user defined relation $\ldbrack \langle e_1,e_2,r\rangle \rdbrack$
$\subseteq \ldbrack e_1 \rdbrack \times \ldbrack e_2 \rdbrack \times \ldbrack r \rdbrack$.

>>>>

Given a word $w\in L$, 
$Sy(w)=\{u\mid u\in \ss \land\exists s\in \gamma(w)\land~\alpha(u)\cap\alpha(s)\neq\emptyset\}$
is the set of Arabic stems 
directly related to $w$ through the gloss map.
%
Let $Sy^{i}(w)$ denote stems related to $w$ using the gloss map of order $i$ recursively such that
$Sy^{1}(w) = Sy(w)$ and
$Sy^{i+1}(w)=\{u\mid u\in S\land\exists s\in Sy^{i}(w)\land~\alpha(u)\cap\alpha(s)\neq\emptyset\}$.
Formally, $Syn^k(w) = \bigcup\limits_{i=1}^{k} Sy^{i}(w)$ for $i\in[1 \ldots k]$.
%
\major{
The example in Figure~\ref{fig:introsynEx} illustrates the computation.
Let $w$ denote an input Arabic word \RL{mA'}, which has the gloss \cci{water}, i.e. \cci{water} $\in\alpha(w)$.
$w$ shares this gloss with the stem \RL{n.d.h}, denoted $s_1$, i.e. $s_1\in Sy^{1}(w)$.
%The input word \RL{mA'} shares the gloss \RL{n.d.h} through the gloss intersection \cci{water}. 
Next, the stem \RL{r^s^s}, denoted $s_2$, shares the gloss \cci{spray} with $s_1$, i.e. $s_2\in Sy^{1}(s1)\subset Sy^{2}(w)$.
Therefore, $Syn^2(w)$ relates the words \RL{mA'} and \RL{r^s^s}.
}

@@

\major{
Consider the task we discussed in the introduction (Figure~\ref{fig:intromotiv}) 
and recall that we are interested in identifying names of people, names of places, relative positions, and numerical terms.
The table in Figure~\ref{fig:taskMBF} presents the defined formulae.
The user denotes the ``name of person'' entities with formula $N$ 
which requires the {\em category} feature in the morphological solution of a word to be {\tt Name\_of\_Person}.
The entities $n_1$, $n_2$, and $n_3$ are matches of the formula $N$ in the text.
Similarly, the user specifies formula $P$ to denote ``name of place'' entities. 
The user specifies formula $R$ to denote ``relative position'' entities, 
and defines it as a disjunction of terms that check for 
solutions matching stems such as \RL{qrb} (near) and \RL{fy} (in).
%requires the stem feature of a word to belong to a selected list of stems containing \RL{fy} and \RL{qrb}. a
Similarly, $U$ denotes numerical terms and is a disjunction of constraints 
requiring the stem feature to belong to a set of stems such as
\RL{'wl}(first), \RL{_tAny}(second), \dots \RL{`A^sr}(tenth).
}

@@

A relation is defined by the user as a tuple 
$\langle e_1,e_2,r\rangle$ where 
$e_1,e_2,$ and $r$ are identifiers associated with 
sub-expressions of an MRE $f$.
Matches of the relation are a set of labeled binary edges
where matches of $e_1$ and $e_2$ are the source and destination nodes
and matches of $r$ are the edge labels.
We denote $\ldbrack \langle e_1,e_2,r\rangle \rdbrack$ to be the set of matches of the corresponding relation, 
and we refer to them as relational entities.

\major{
We are interested in constructing the relational entity graph in Figure~\ref{fig:intromotiv}.
Let $e_1,o_1,r,o_2,$ and $e_2$ be identifiers to the sub-expressions 
$(P|N)+$,$O?$,$R$,$O\wedge 2$, and $(P|N|U)+$, respectively.
%We select $(P|N)+$ to denote an entity as it captures a composite place name such as \cci{Khalifa tower}.
The matches to $e_1$, $r$, $o_2$, and $e_2$ in match 1 (Fig.~\ref{fig:taskMRE}) are \RL{brj _hlyfT} (Khalifa Tower), \RL{bAlqrb} (next), \RL{mn} (to), and \RL{AltqA.t` Al-'wl} (first intersection).
Note that there is no match to the optional $O$ formula in match 1.
Similarly, the matches to $e_1$, $o_1$, $r$, $o_2$, and $e_2$ in the second \major{matching parse tree} are
\RL{dby mwl} (Dubai Mall), \RL{`l_A} (is located), \RL{mqrbT} (near), \RL{mn h_dA} (this), and \RL{Almbn_A} (building), respectively.
}

\major{
We define the semantic relations $\langle e_1, e_2, r\rangle$, $\langle r, e_1, o_1\rangle$, and $\langle r,e_2,o_2\rangle$.
Relation $\langle e_1,e_2,r\rangle$ creates the 
edge labeled \cci{next to} between \cci{Khalifa tower} and 
\cci{intersection 1} nodes from match 1, and the 
edge labeled \cci{near} between \cci{Dubai Mall} and \cci{the building} nodes from match 2.
Relation $\langle r,e_1,o_1\rangle$ creates the edge labeled \cci{prep} 
between \cci{Dubai Mall} and \cci{near} nodes from match 2. 
Relation $\langle r,e_2,o_2\rangle$ creates the edge labeled \cci{from} 
between \cci{intersection 1} and \cci{next to} nodes in 
match 1, and the 
edge labeled \cci{from this} between \cci{near} and \cci{the building} 
nodes in match 2.
}

@@

For each word $t_i\in T, 0\le i < |T|$.
\framework computes a Boolean value 
for all MBFs. 
\major{
For example, \RL{brj} matches MBF $P$. 
}
Then, it computes the set of MBF tags
$R_i=\{(t_i,tt)| tt=\langle l,f,d\rangle \wedge
f~\mathit{is~an~MBF} \wedge f(t_i)\} \subseteq T \times {\cal T}$
which tags a word $t_i$ with $\mathit{tt}$ 
iff the MBF $f$ associated with
tag type $\mathit{tt}$ is true for $t_i$. 
The MBF evaluation results in a sequence of tag sets 
$\langle R_0, R_1, \ldots, R_{n-1}\rangle$.
If a word $t_i$ has no tag type match, 
its tag set $R_i$ is by default the singleton $O=\{\mathit{NONE}\}$.
\major{
For example, the tag sets for the text in Figure~\ref{f:overview} follows 
$\{ \{NONE\}, \{NONE\}, \{NONE\}, \{NONE\}, \{($\RL{brj}$,P)\}, \{($\RL{xlyfT}$,N)\},\ldots\}$. 
}

@@

\framework simulates the generated NFA over the sequence of tag sets matching the MBF formulae.
%A simulation match $m$ of an expression $f$ is a tree where the root is the MRE expression, the internal nodes are the MRE and MBF operations, and the leaves are matches of the MAT terms of $f$.
A simulation match $m$ of an expression $f$ is a \major{parse tree} where the root spans the expression, the internal nodes are roots to sub-expressions of $f$, and the leaves are matches of the MBF formulae of $f$, e.g. Fig.~\ref{fig:taskMRE}.
The sequence of leaf matches forms a vector of tags $\langle r_k,r_{k+1},\dots,r_j\rangle$ 
corresponding to the text sequence 
$\langle t_k,t_{k+1},\dots,t_j\rangle$ where $r_{\ell}\in R_{\ell},0\le k\le \ell \le j < n$.
%
%If the simulation has a match $\langle r_m,r_{m+1},\dots,r_n\rangle$ 
%where $0\le m\le n$, 
%the next simulation starts at $R_{n+1}$. 
%This disallows overlap of matches for the same MRE. 
%
%In case the NFA simulation has no match, 
%the next simulation starts at $R_{m+1}$. 
If we have more than one match for an expression, 
%starting at $R_k$ where $0\le k\le n$, 
\framework returns the longest.
%Figure~\ref{fig:taskMRE} shows two match trees of $\mathit{dir}$ extracted from the text of Figure~\ref{fig:intromotiv}.
%\RL{dby} and \RL{mwl} are leaf nodes referring to name of place tags ($P$).
%The $+$, $|$, and $?$ MRE operations are internal nodes.

@@

%\begin{figure}[tb!]
%{ \relsize{-1}
%\begin{framed}
%\begin{multicols}{2}
%\begin{itemize}
%\item Annotated Expression: \\
%\\
%{ \relsize{-2.5}
%%\begin{itemize}
%%  \item 
%  $\stackrel{e_1}{(P|N)+}~\stackrel{o_1}{\mathit{O}?}~\stackrel{r}{R}~\stackrel{o_2}{\mathit{O^{\wedge}2}}~\stackrel{e_2}{(P|N|U)+}$
%%\end{itemize}
%}
%\end{itemize}
%\begin{itemize}
%\item User defined semantic \\relations:
%\begin{itemize}
%\item $\langle e_1,e_2,r\rangle$
%\item $\langle r,e_1,o_1\rangle$
%\item $\langle r,e_2,o_2\rangle$
%\end{itemize}
%\end{itemize}
%\columnbreak
%\setcode{utf8}
%\transfalse
%\resizebox{0.9\columnwidth}{!}{
%	\relsize{+1.5} \input{figures/ergraphEx.tex}
%}
%\transtrue
%\setcode{standard}
%\end{multicols}
%\end{framed}
%}
%\caption{User-defined semantic relation example}
%\label{fig:srEx}
%\end{figure}
%
Finally, \framework computes the relational entities corresponding to each user defined relation $\ldbrack \langle e_1,e_2,r\rangle \rdbrack$
$\subseteq \ldbrack e_1 \rdbrack \times \ldbrack e_2 \rdbrack \times \ldbrack r \rdbrack$.
